---
title: "Characteristics between diabetic and non-diabetic groups"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Chi-square test for categorical variables
```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(knitr)
chi_diabetes = read.csv("Diabete datasets/diabetes_binary_health_indicators_BRFSS2015.csv")

cat_vars <- c(
  "HighBP", "HighChol", "CholCheck","Smoker", "Stroke", "HeartDiseaseorAttack",
  "PhysActivity", "Fruits", "Veggies", "HvyAlcoholConsump",
  "AnyHealthcare", "NoDocbcCost", "DiffWalk",
  "Sex", "Age", "Education", "Income"
)

chi_results <- list()

for (v in cat_vars) {
  tbl <- table(chi_diabetes$Diabetes_binary, chi_diabetes[[v]])
  test <- chisq.test(tbl)
  chi_results[[v]] <- test
}

chi_table <- data.frame(
  Variable   = names(chi_results),
  Chi_square = sapply(chi_results, function(x) unname(x$statistic)),
  df         = sapply(chi_results, function(x) unname(x$parameter)),
  p_value    = sapply(chi_results, function(x) x$p.value)
)

chi_table$p_value <- format(chi_table$p_value, scientific = TRUE) #show detailed p-value result

kable(chi_table, digits = 4, row.names = FALSE, caption = "Chi-square test results for categorical variables") #put everything neatly in a table
```

## Chi-square test result
**All categorical variables** showed statistically significant differences between diabetic and non-diabetic participants (all p < 0.001).

# T-test for continuous variables
```{r t_test, warning=FALSE, message=FALSE}
cont_vars <- c("BMI", "MentHlth", "PhysHlth", "GenHlth")

# Run t-tests and extract key results
t_results <- lapply(cont_vars, function(v) {
  
  tt <- t.test(chi_diabetes[[v]] ~ chi_diabetes$Diabetes_binary)
  
  data.frame(
    Variable        = v,
    Mean_non_diab   = unname(tt$estimate[1]),
    Mean_diab       = unname(tt$estimate[2]),
    t_statistic     = unname(tt$statistic),
    df              = unname(tt$parameter),
    p_value         = tt$p.value
  )
})

# Combine list rows into a data frame
t_table <- bind_rows(t_results)

# Format values
t_table <- t_table |>
  mutate(
    Mean_non_diab = round(Mean_non_diab, 2),
    Mean_diab     = round(Mean_diab, 2),
    t_statistic   = round(t_statistic, 2),
    df            = round(df, 1),
    p_value       = p_value)

t_table$p_value <- formatC(t_table$p_value, format = "e", digits = 2)

# Print table
kable(t_table, caption = "Summary of t-tests")
```

## T-test result
All tests showed **highly significant differences** (all p < 2.2 × 10⁻¹⁶), indicating that the mean health outcomes differ substantially between the two groups.


# Visualization for continuous variables
```{r t_test_plot, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)

df_cont <- 
  chi_diabetes |>
  pivot_longer(cols = all_of(cont_vars),
               names_to = "variable",
               values_to = "value")

ggplot(df_cont,aes(x = factor(Diabetes_binary,
                    levels = c(0, 1),
                    labels = c("Non-diabetes", "Diabetes")),
                    y = value,
                    fill = factor(Diabetes_binary))) +
  geom_boxplot(outlier.alpha = 0.15) +
  facet_wrap(~ variable, scales = "free_y") +
  labs(
    title = "Continuous variables of Diabetes Status",
    x = "Diabetes Status",
    y = "Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## Comments on the plot

- **BMI:** Diabetic individuals display noticeably higher BMI levels.
- **GenHlth:** Diabetic participants report worse general health compared to the non-diabetic group.
- **MentHlth:** Diabetic individuals experience more days of poor mental health.
- **PhysHlth:** The diabetic group reports substantially more days of poor physical health.