---
title: "Data Overview"
output: 
  html_document:
    toc: true
    code_folding: hide
    toc_float: true
    theme: cosmo
---
### <b>Data Source</b>
This Kaggle dataset [Diabetes Health Indicators](https://www.kaggle.com/datasets/alexteboul/diabetes-health-indicators-dataset?select=diabetes_binary_5050split_health_indicators_BRFSS2015.csv) used data collected from Behavioral Risk Factor Surveillance System (BRFSS), a health-related telephone survey that is collected annually by the CDC. Each year, the survey collects responses from over 400,000 Americans on health-related risk behaviors, chronic health conditions, and the use of preventative services. It has been conducted every year since 1984. This specific dataset was from 2015. These features are either questions directly asked of participants, or calculated variables based on individual participant responses.

### <b>Data Cleaning</b>

```{r, include = FALSE, message = FALSE, warning = FALSE}
# Load required libraries
library(tidyverse)
library(rvest)
library(plotly)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```


```{r}
#Load & Clean Dataset
diabetes = 
  read_csv("Diabete datasets/diabetes_012_health_indicators_BRFSS2015.csv") |>
  janitor::clean_names() |> 
  mutate(
    sex = case_match(
      sex,
      0 ~ "Female",
      1 ~ "Male"
    ),
    sex = factor(sex),
    age = case_match(
      as.numeric(age),
      1 ~ "18-24",
      2 ~ "25–29",
      3 ~ "30–34",
      4 ~ "35–39",
      5 ~ "40–44",
      6 ~ "45-49",
      7 ~ "50-54",
      8 ~ "55-59",
      9 ~ "60-64",
      10 ~ "65-69",
      11 ~ "70-74",
      12 ~ "75-79",
      13 ~ "80+"
    ),
    age = 
      factor(age, 
             levels = 
               c("18-24","25–29", "30–34", "35–39",
                 "40–44","45-49","50-54","55-59",
                 "60-64","65-69","70-74","75-79","80+"), 
             ordered = TRUE),
    education = case_match(
      as.numeric(education),
      1 ~ "Never attended school",
      2 ~ "Grade 1-8",
      3 ~ "Grade 9-11",
      4 ~ "High school graduate",
      5 ~ "College 1-3",
      6 ~ "College graduate"
     ),
    education = 
      factor(education, 
             levels = 
               c("Never attended school or only kindergarten ","Grade 1-8", "Grade 9-11",
                 "High school graduate","College 1-3",
                 "College graduate"), 
             ordered = TRUE),
    income = case_match(
      as.numeric(income),
      1 ~ "Less than $10,000",
      2 ~ "$10,000 - $15,000",
      3 ~ "$15,000 - $20,000",
      4 ~ "$20,000 - $25,000",
      5 ~ "$25,000 - $35,000",
      6 ~ "$35,000 - $50,000",
      7 ~ "$50,000 - $75,000",
      8 ~ "$75,000 +"
    ),
    income = 
      factor(income, 
             levels = 
               c("Less than $10,000","$10,000 - $15,000",
                 "$15,000 - $20,000","$20,000 - $25,000",
                 "$25,000 - $35,000","$35,000 - $50,000",
                 "$50,000 - $75,000","$75,000 +"), 
             ordered = TRUE),
    gen_hlth = case_match(
      as.numeric(gen_hlth),
      1 ~ "Poor",
      2 ~ "Fair",
      3 ~ "Good",
      4 ~ "Very Good",
      5 ~ "Excellent"
    ),
    gen_hlth = 
      factor(gen_hlth, 
             levels = 
               c("Poor", "Fair", "Good", "Very Good", "Excellent"), 
             ordered = TRUE),
    diabetes_012 = 
      factor(diabetes_012,
             levels = c(0,1,2),
             labels = c("No Diabetes", "Prediabetes", "Diabetes")),
    diabetes_2 = case_when(
      diabetes_012 == "No Diabetes" ~ 0,
      diabetes_012 != "No Diabetes" ~ 1
    ),
    diabetes_2 = as.factor(diabetes_2)
  )
  
binary_vars = c(
  "high_bp", "high_chol", "chol_check",
  "smoker", "stroke", "heart_diseaseor_attack",
  "phys_activity", "hvy_alcohol_consump", "any_healthcare",
  "no_docbc_cost", "diff_walk", "fruits", "veggies"
)

diabetes[binary_vars] = lapply(
  diabetes[binary_vars],
  function(x) factor(ifelse(x == 1, "Yes", "No"))
)

```


Data have `r nrow(diabetes)` records and `r ncol(diabetes)` variables.
Each record corresponds to an individual respondent and includes information on demographics, health conditions, lifestyle behaviors, and healthcare access. 

In our data cleaning and preprocessing steps, we did the following:

#### 1. Converted numeric codes into meaningful categorical labels

For example, the sex variable was coded as 0 and 1. We converted it to "Female" and "Male" and then turned it into a factor variable.

Similarly, age, income, GenHlth(general health) and education were originally numeric codes representing ranges or categories. We mapped these codes to descriptive labels and and specified ordered factors so that analyses respect the natural order of these categories.

#### 2. Encoded binary health indicators as factors

Several binary variables, such as high_bp (high blood pressure), smoker, and phys_activity (physical activity), were coded as 0 and 1.

We converted these to "Yes"/"No" factors to make the dataset more interpretable.

#### 3. Created a categorical target variable for diabetes status

The variable `diabetes_012` indicates whether a person has No Diabetes, Prediabetes, or Diabetes. This is a factor with three levels. For some analyses, we also created a binary target variable, `diabetes_2`, which groups individuals into two categories: non-diabetes and diabetes.

This binary variable simplifies modeling to predict the presence or absence of diabetes, while the original three-level factor can still be used for more detailed analyses or exploratory data visualization.


### <b>Data Summary</b>
```{r}
#data summary
library(skimr)
skim(diabetes)
```


### <b>Data Variables</b>
The data variables that will be included in the EDA are:

`Diabetes_012`: 0 = no diabetes 1 = prediabetes 2 = diabetes

`Diabetes_2` : 0 = no diabetes 1 = prediabetes/diabetes

`HighBP`: high blood pressure (0 = no 1 = yes)

`HighChol`: high cholesterol (0 = no 1 = yes)

`CholCheck`: cholesterol check in 5 years (0 = no 1 = yes)

`BMI`: Body Mass Index

`Smoker`: Have you smoked at least 100 cigarettes in your entire life? (0 = no 1 = yes)

* [Note: 5 packs = 100 cigarettes] 

`Stroke`: (ever told) you had a stroke (0 = no 1 = yes)

`HeartDiseaseorAttack`: coronary heart disease (CHD) or myocardial infarction (MI) (0 = no 1 = yes)

`PhysActivity`: physical activity in past 30 days (0 = no 1 = yes)

* not including job 

`Fruits`: consume fruit 1 or more times per day (0 = no 1 = yes)

`Veggies`: consume vegetables 1 or more times per day (0 = no 1 = yes)

`HvyAlcoholConsump`: adult men >=14 drinks per week and adult women>=7 drinks per week (0 = no 1 = yes)

`AnyHealthcare`: have any kind of health care coverage, including health insurance, prepaid plans such as HMO, etc. (0 = no 1 = yes)

`NoDocbcCost`: was there a time in the past 12 months when you needed to see a doctor but could not because of cost? (0 = no 1 = yes)

`GenHlth`: would you say that in general your health is: scale 1-5 

* 1 = excellent 
* 2 = very good 
* 3 = good 
* 4 = fair 
* 5 = poor

`MentHlth`: which includes stress, depression, and problems with emotions, for how many days during the past 30 days was your mental health not good?

* scale 1-30 days

`PhysHlth`: which includes physical illness and injury, for how many days during the past 30 days was your physical health not good?

* scale 1-30 days

`DiffWalk`: do you have serious difficulty walking or climbing stairs? (0 = no 1 = yes)

`Sex`: 0 = female 1 = male

`Age`: 13-level age category 

* 1 = 18-24
* 2 = 25–29
* ...
* 9 = 60-64 
* 13 = 80 or older

`Education`: education level: scale 1-6 

* 1 = Never attended school or only kindergarten 
* 2 = Grade 1-8
* 3 = Grade 9-11
* 4 = High school graduate
* 5 = College 1-3
* 6 = College graduate

`Income`: income scale: scale 1-8 

* 1 = less than $10,000
* 2 = $10,000 - $15,000
* 3 = $15,000 - $20,000
* 4 = $20,000 - $25,000
* ...
* 8 = $75,000 +

