---
title: "Exploring Diabetes Risk Factors"
output:
  html_document:
    self_contained: false
    toc: true
    toc_float: true
    theme: cosmo

---

## <b>Diabetes distribution</b>

```{r, echo=FALSE}
diabetes |>  
  count(diabetes_012) |>  
  plot_ly(
    x = ~diabetes_012,
    y = ~n,
    type = "bar",
    color = ~diabetes_012
  ) |>  
  layout(
    barmode = "stack",
    title = "Diabetes Prevalence",
    xaxis = list(title = "Age Group"),
    yaxis = list(title = "Count")
  )
```

## <b>Sex & Age distribution</b>

```{r , echo=FALSE}
diabetes |>
  count(sex, age) |>  
  plot_ly(
    x = ~age,
    y = ~n,
    type = "bar",
    color = ~sex,
    colors = c("tomato","skyblue")
  ) %>% 
  layout(
    barmode = "group",
    title = "Sex and Age Group Distribution",
    xaxis = list(title = "Age Group"),
    yaxis = list(title = "Count")
  )
```

```{r , echo=FALSE}
diabetes |>
  group_by(age, sex) |> 
  summarise(diabetes_rate = mean(diabetes_012 == "Diabetes")) |>  
  plot_ly(
    x = ~age,
    y = ~diabetes_rate,
    color = ~sex,
    colors = c("tomato","skyblue"),
    type = "scatter",
    mode = "lines+markers"
  ) |>  
  layout(
    title = "Diabetes Prevalence by Sex and Age Group",
    xaxis = list(title = "Age Group"),
    yaxis = list(title = "Diabetes Rate")
  )
```


## <b>BMI Distribution</b>

```{r , echo=FALSE}
plot_ly(
  data = diabetes,
  x = ~diabetes_012,
  y = ~bmi,
  color = ~diabetes_012,
  type = "box"
  ) |> 
  layout(
    title = "BMI Distribution by Diabetes Status",
    xaxis = list(title = "Diabetes Status"),
    yaxis = list(title = "BMI")
  )
```



```{r, echo=FALSE}
# filtering prediabetes
diabetes_nd = diabetes |>
  filter(diabetes_012 != "Prediabetes") |>
  droplevels()

# long format
binary_long_nd = diabetes_nd |>
  select(diabetes_012, all_of(binary_vars)) |>
  pivot_longer(
    cols = all_of(binary_vars),
    names_to = "variable",
    values_to = "value"
  )

# proportion "Yes" in each group
binary_yes_nd = binary_long_nd |>
  filter(value == "Yes") |>
  count(variable, diabetes_012) |>
  group_by(variable) |>
  mutate(prop = n / sum(n)) |>
  ungroup()

# proportion "No" in each group
binary_no_nd = binary_long_nd |>
  filter(value == "No") |>
  count(variable, diabetes_012) |>
  group_by(variable) |>
  mutate(prop = n / sum(n)) |>
  ungroup()
```

## <b>Binary Variables</b>

We started by splitting up the variables into three categories: binary variables, continuous variables and ordered categorical variables. Then, we compared all the variables within each category to diabetes. The binary variables are separated into three main parts: clinical conditions, lifestyle, and access to healthcare / function. 

### <b>Clinical Conditions vs Diabetes</b>
Clinical conditions include high blood pressure, high cholesterol, stroke, and a history of heart disease or heart attack. The bar chart shows the proportions of participants who reported each condition and whether they also have diabetes. 
```{r, echo=FALSE}
clinical_vars = c("high_bp", "high_chol", "stroke", "heart_diseaseor_attack")

# answered yes
p_clinical = binary_yes_nd |>
  filter(variable %in% clinical_vars) |>
  ggplot(aes(x = variable, y = prop, fill = diabetes_012)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Clinical Conditions: Proportion 'Yes' by Diabetes Status",
    x = NULL,
    y = "Proportion 'Yes'",
    fill = "Diabetes status"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))


ggplotly(p_clinical)

```

Among those who answered 'yes,' a history of heart disease or heart attack has the highest proportion of diabetes at 33.9% (~0.339). In contrast, participants with high cholesterol have the highest proportion of not having diabetes.


```{r, echo=FALSE}
# answered no
p_clinical_no = binary_no_nd |>
  filter(variable %in% clinical_vars) |>
  ggplot(aes(x = variable, y = prop, fill = diabetes_012)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Clinical Variables: Proportion 'No' by Diabetes Status",
    x = NULL,
    y = "Proportion 'No'",
    fill = "Diabetes status"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

ggplotly(p_clinical_no)
```

This shows among those who answered 'no' to the clinical conditions, people who don’t have high blood pressure have the lowest proportion to have diabetes (~0.061). On the other hand, people who don’t have strokes have the highest proportion of having diabetes (~0.134).


### <b>Lifestyle vs Diabetes</b>
The next category is lifestyle which include smoking, physical activity, fruits and vegetables intake, and heavy alcohol consumption. 
```{r, echo=FALSE}
behavior_vars = c("smoker", "phys_activity", "fruits", "veggies", "hvy_alcohol_consump")

# answered yes
p_behavior = binary_yes_nd |>
  filter(variable %in% behavior_vars) |>
  ggplot(aes(x = variable, y = prop, fill = diabetes_012)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Lifestyle Behaviors: Proportion 'Yes' by Diabetes Status",
    x = NULL,
    y = "Proportion 'Yes'",
    fill = "Diabetes status"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))


ggplotly(p_behavior)

```

Among those who answered 'yes' to these conditions, people who smoke have the highest proportion (~0.166) of having diabetes compared to rest variables. In contrast, people who have heavy alcohol consumption have the lowest proportion of people with diabetes (~0.059).

```{r, echo=FALSE}

# answered no
p_behavior_no = binary_no_nd |>
  filter(variable %in% behavior_vars) |>
  ggplot(aes(x = variable, y = prop, fill = diabetes_012)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Lifestyle Behaviors: Proportion 'No' by Diabetes Status",
    x = NULL,
    y = "Proportion 'No'",
    fill = "Diabetes status"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

ggplotly(p_behavior_no)

```
This shows among those who answered 'no' to the lifestyle habits, people who don’t do physical activities have the highest proportion of having diabetes (~0.217). Those who don't smoke have the lower proportion of having diabetes (~0.123).

### <b>Healthcare Access and Function vs Diabetes</b>
Lastly, the variables in healthcare access and function category include cholesterol check, health care coverage, health care affordability and difficulty walking.
```{r, echo=FALSE}
access_vars = c("chol_check", "any_healthcare", "no_docbc_cost", "diff_walk")

p_access = binary_yes_nd |>
  filter(variable %in% access_vars) |>
  ggplot(aes(x = variable, y = prop, fill = diabetes_012)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Healthcare Access and Function: Proportion 'Yes' by Diabetes Status",
    x = NULL,
    y = "Proportion 'Yes'",
    fill = "Diabetes status"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

ggplotly(p_access)

```

Among participants who answered ‘yes’ to these conditions, those who have difficulty walking have the highest proportion (~0.317) of having diabetes compared to the other variables. Those who have healthcare have the lowest proportion of having diabetes (~0.143).



```{r, echo=FALSE}
# answered no
p_access_no = binary_no_nd |>
  filter(variable %in% access_vars) |>
  ggplot(aes(x = variable, y = prop, fill = diabetes_012)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Healthcare Access and Function: Proportion 'No' by Diabetes Status",
    x = NULL,
    y = "Proportion 'No'",
    fill = "Diabetes status"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

ggplotly(p_access_no)
```
Among the participants who answered 'no,'those who have not experienced healthcare cost affordability issue have the highest proportion of having diabetes (~0.138). On the other hand, those who have not had a cholesterol check in the past five years have the lowest diabetes proportion (~0.026).
