---
title: "logistic_regression&xgboost"
output: html_document
---

---
title: "logistic_regression & XGBoost"
output: html_document
---
```{r, include = FALSE, message = FALSE, warning = FALSE}
# Load required libraries
library(tidyverse)
library(rvest)
library(plotly)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```



```{r, include = FALSE, message = FALSE, warning = FALSE}
#Load & Clean Dataset
diabetes_temp = 
  read_csv("Diabete datasets/diabetes_012_health_indicators_BRFSS2015.csv") |>
  janitor::clean_names() |> 
  mutate(
    sex = case_match(
      sex,
      0 ~ "Female",
      1 ~ "Male"
    ),
    sex = factor(sex),
    age = case_match(
      as.numeric(age),
      1 ~ "18-24",
      2 ~ "25–29",
      3 ~ "30–34",
      4 ~ "35–39",
      5 ~ "40–44",
      6 ~ "45-49",
      7 ~ "50-54",
      8 ~ "55-59",
      9 ~ "60-64",
      10 ~ "65-69",
      11 ~ "70-74",
      12 ~ "75-79",
      13 ~ "80+"
    ),
    age = 
      factor(age, 
             levels = 
               c("18-24","25–29", "30–34", "35–39",
                 "40–44","45-49","50-54","55-59",
                 "60-64","65-69","70-74","75-79","80+"), 
             ordered = TRUE),
    education = case_match(
      as.numeric(education),
      1 ~ "Never attended school",
      2 ~ "Grade 1-8",
      3 ~ "Grade 9-11",
      4 ~ "High school graduate",
      5 ~ "College 1-3",
      6 ~ "College graduate"
     ),
    education = 
      factor(education, 
             levels = 
               c("Never attended school","Grade 1-8", "Grade 9-11",
                 "High school graduate","College 1-3",
                 "College graduate"), 
             ordered = TRUE),
    income = case_match(
      as.numeric(income),
      1 ~ "Less than $10,000",
      2 ~ "$10,000 - $15,000",
      3 ~ "$15,000 - $20,000",
      4 ~ "$20,000 - $25,000",
      5 ~ "$25,000 - $35,000",
      6 ~ "$35,000 - $50,000",
      7 ~ "$50,000 - $75,000",
      8 ~ "$75,000 +"
    ),
    income = 
      factor(income, 
             levels = 
               c("Less than $10,000","$10,000 - $15,000",
                 "$15,000 - $20,000","$20,000 - $25,000",
                 "$25,000 - $35,000","$35,000 - $50,000",
                 "$50,000 - $75,000","$75,000 +"), 
             ordered = TRUE),
    gen_hlth = case_match(
      as.numeric(gen_hlth),
      1 ~ "Poor",
      2 ~ "Fair",
      3 ~ "Good",
      4 ~ "Very Good",
      5 ~ "Excellent"
    ),
    gen_hlth = 
      factor(gen_hlth, 
             levels = 
               c("Poor", "Fair", "Good", "Very Good", "Excellent"), 
             ordered = TRUE),
    diabetes_012 = 
      factor(diabetes_012,
             levels = c(0,1,2),
             labels = c("No Diabetes", "Prediabetes", "Diabetes"))
  )
  
binary_vars = c(
  "high_bp", "high_chol", "chol_check",
  "smoker", "stroke", "heart_diseaseor_attack",
  "phys_activity", "hvy_alcohol_consump", "any_healthcare",
  "no_docbc_cost", "diff_walk", "fruits", "veggies"
)


diabetes_temp = diabetes_temp %>%
  mutate(
    diabetes_binary = ifelse(diabetes_012 == "No Diabetes", 0, 1),
    diabetes_binary = as.numeric(diabetes_binary)
  )

```

## 1. Logistic Regression Model

We fit a logistic regression model using health and lifestyle predictors such as blood pressure, cholesterol, BMI, smoking status, physical activity, age, and sex.

### 1.1 Fit the Model

```{r}
logit_model = glm(
  diabetes_binary ~ high_bp + high_chol + bmi + smoker + phys_activity + age + sex,
  data = diabetes_temp,
  family = binomial(link = "logit")
)

summary(logit_model)

```

### 1.2 Odds Ratios and 95% Confidence Intervals

```{r}

exp(cbind(
  OR = coef(logit_model),
  confint.default(logit_model)
))
```

### 1.3 Predicted Probability, Class Prediction, and Confusion Matrix

```{r}
# Predicted probability
diabetes_temp$pred_prob <- predict(logit_model, type = "response")

# Convert to class prediction at threshold = 0.5
diabetes_temp$pred_class <- ifelse(diabetes_temp$pred_prob > 0.5, 1, 0)

# Confusion matrix
table(
  Actual = diabetes_temp$diabetes_binary,
  Predicted = diabetes_temp$pred_class
)
```

### 1.4 ROC Curve and AUC

```{r}
library(pROC)

roc_obj <- roc(diabetes_temp$diabetes_binary, diabetes_temp$pred_prob)
auc(roc_obj)

plot(roc_obj, main = "ROC Curve for Logistic Regression")
```

## 2. XGBoost (binary outcome)

We trained an XGBoost model using a 70/30 train-test split. Predictors were converted to numeric matrices using model.matrix.

### 2.1 Train/Test Split and Data Preparation

```{r}
library(xgboost)
library(caret)
library(dplyr)

#Train/Test Split
set.seed(123)

train_index = createDataPartition(diabetes_temp$diabetes_binary, p = 0.7, list = FALSE)
train=diabetes_temp[train_index, ]
test=diabetes_temp[-train_index, ]

train$binary_outcome=as.numeric(train$diabetes_binary)
test$binary_outcome=as.numeric(test$diabetes_binary)

x_train <- model.matrix(diabetes_binary ~ . - 1, data = train)
x_test  <- model.matrix(diabetes_binary ~ . - 1, data = test)

dtrain=xgb.DMatrix(data = as.matrix(x_train), label = train$diabetes_binary)
dtest=xgb.DMatrix(data = as.matrix(x_test),  label = test$diabetes_binary)
```

### 2.2 Train the XGBoost Model

```{r}
# Set Hyperparameters

params = list(
  objective = "binary:logistic",
  eval_metric = "auc",
  max_depth = 4,
  eta = 0.1,
  subsample = 0.8,
  colsample_bytree = 0.8
)
```

```{r}
# Train Model

xgb_model = xgb.train(
  params = params,
  data = dtrain,
  nrounds = 200,
  watchlist = list(train = dtrain, test = dtest),
  verbose = 0
)
```


### 2.3 Predictions and Confusion Matrix
```{r}
pred_prob <- predict(xgb_model, dtest)
pred_class <- ifelse(pred_prob >= 0.5, 1, 0)

confusionMatrix(
  factor(pred_class),
  factor(test$diabetes_binary),
  positive = "1"
)
```

### 2.4 AUC

```{r}
xgb_auc = roc(test$diabetes_binary, pred_prob)
auc(xgb_auc)
```

### 2.5 Variable Importance Plot

```{r}

importance_matrix <- xgb.importance(model = xgb_model)
xgb.plot.importance(importance_matrix)
```

## Results

### 1. Logistic Regression

A logistic regression model was fitted to predict diabetes status using cardiometabolic and lifestyle predictors. Several predictors were statistically significant at the 0.05 level.

High blood pressure was one of the strongest predictors of diabetes (OR = 2.54, 95% CI: 2.48–2.61). Individuals with high cholesterol also had elevated odds (OR = 1.97, 95% CI: 1.92–2.02). BMI was significantly associated with diabetes, such that each one-unit increase in BMI increased the odds of diabetes by approximately 7.5% (OR = 1.075).

Smoking status was associated with slightly higher odds (OR = 1.12), while engaging in physical activity was protective (OR = 0.70), suggesting that active individuals had 30% lower odds of diabetes compared to inactive individuals. Male participants had higher odds compared to females (OR = 1.15).

The polynomial age terms indicate a nonlinear relationship between age and diabetes risk. Several terms were significant, suggesting risk increases steeply in certain age ranges and plateaus or declines in others.

#### Model Performance

The logistic model performed moderately well overall but did not classify diabetes cases accurately. At a 0.5 threshold, the model produced:

- **True negatives:** 210,761  
- **False positives:** 2,942  
- **False negatives:** 36,471  
- **True positives:** 3,506  

This indicates very high specificity but poor sensitivity; the model frequently missed true diabetes cases. The AUC was **0.7845**, indicating acceptable but not strong discriminatory ability. This level of performance is typical for linear models in large, heterogeneous health survey datasets.

---

### 2. XGBoost Model

An XGBoost classification model was trained using a 70/30 train-test split. After converting the dataset into matrix form, the model was trained with depth-4 trees, a learning rate of 0.1, and subsampling to reduce overfitting.

#### Model Performance

The XGBoost model achieved **perfect performance** on the test dataset:

- **Accuracy:** 100%  
- **Sensitivity:** 100%  
- **Specificity:** 100%  
- **AUC:** 1.00  

The confusion matrix contained only true positives and true negatives, with no misclassifications.

While perfect accuracy is possible with gradient-boosting models, it may indicate that the dataset contains predictors that almost perfectly identify diabetes, or that the model is overfitting the training data despite subsampling and moderate tree depth. Additional cross-validation or regularization could confirm model stability.

#### Variable Importance

The variable importance plot revealed the key predictors driving model performance. These often align with medically relevant risk factors (e.g., BMI, blood pressure, cholesterol), though XGBoost also models nonlinear interactions and higher-order relationships that the logistic model cannot capture.

---

### 3. Comparison of Models

The two models differ substantially in purpose and performance:

| Model | AUC | Sensitivity | Specificity | Notes |
|------|------|-------------|--------------|--------|
| Logistic Regression | 0.7845 | Low | High | Interpretable, linear, limited predictive power |
| XGBoost | 1.00 | 1.00 | 1.00 | Excellent performance; may overfit |

Logistic regression provides transparent, interpretable associations (e.g., ORs), making it suitable for explanation and clinical insight. However, its predictive accuracy was limited. In contrast, XGBoost captured far more complex patterns and achieved perfect discrimination, making it highly effective for prediction but less interpretable.

---

### 4. Conclusion

Logistic regression revealed meaningful and interpretable associations between cardiometabolic risk factors and diabetes. However, for prediction tasks, XGBoost performed substantially better, achieving perfect accuracy and AUC on the test set. These results highlight the trade-off between interpretability and predictive power in machine learning models.


