---
title: "Building Diabetes Prediction Models"
output:
  pdf_document:
    toc: false
  html_document:
    self_contained: false
    theme: cosmo
    toc: false
---

---
```{r, include = FALSE, message = FALSE, warning = FALSE}
library(ggplot2)
library(tidyverse)
library(caret)
library(nnet)
library(dplyr)
library(pROC) 
library(ranger)
```


```{r}
diabetes = readRDS("clean_diabetes_data.rds")
```

## Model 1 Preparation 
```{r}
df_multilog <- diabetes |>
  select(diabetes_012, high_bp, high_chol, chol_check, bmi, smoker, stroke, 
         heart_diseaseor_attack, phys_activity, fruits, veggies, hvy_alcohol_consump,
         any_healthcare, no_docbc_cost, gen_hlth, ment_hlth, phys_hlth, diff_walk, 
         sex, age, education, income)
df_multilog
```

### train/test split
```{r}
# using createDataPartition because diabetes_012 is unbalanced based on EDA.
train_test_split <- createDataPartition(df_multilog$diabetes_012, p = 0.8, list = FALSE)
train <- df_multilog[train_test_split, ]
test  <- df_multilog[-train_test_split, ]

y_test <- test$diabetes_012
```

### F1 function for multiclass
```{r}
multi_f1 <- function(true, pred) {
  # Convert to factors
  true <- as.factor(true)
  pred <- as.factor(pred)
  
  # make sure all classes are taken into consideration (bc diabetes=1 is rare)
  class <- union(levels(true), levels(pred))
  true <- factor(true, levels = class)
  pred <- factor(pred, levels = class)
  
  # A vector to store F1 for each class
  multi_f1_vec <- numeric(length(class))
  
  # Loop over each class
  for (j in seq_along(class)) {
    i <- class[j]               
    tp <- sum(true == i & pred == i)
    fn <- sum(true == i & pred != i)
    fp <- sum(true != i & pred == i)
    rec  <- if((tp + fn) == 0) 0 else tp / (tp + fn)
    prec <- if((tp + fp) == 0) 0 else tp / (tp + fp)
    
    if(prec + rec == 0) {
      multi_f1_vec[j] <- 0
    } else {
      multi_f1_vec[j] <- 2 * prec * rec / (prec + rec)
    }
  }
    mean(multi_f1_vec)
}
```



## Model 1: Multinomial Logistic Regression
```{r}
model1 <- multinom(diabetes_012 ~ ., data = train, trace = FALSE)
```

```{r}
model1_pred <- predict(model1, newdata = test, type = "class")
model1_prob <- predict(model1, newdata = test, type = "prob")

model1_accuracy <- mean(model1_pred == y_test)
cat("Accuracy is:", model1_accuracy, "\n")

model1_macro_f1  <- multi_f1(y_test, model1_pred)
cat("Macro F1 score is:", model1_macro_f1, "\n")

model1__macro_auc <- as.numeric(multiclass.roc(response = y_test, predictor = model1_prob)$auc)
cat("Macro AUC is:", model1__macro_auc, "\n")
```

### 1.1 ROC
```{r}
cols <- c("blue", "green", "pink")

# Compute one-vs-rest ROC and AUC for each class
model1_roc1 <- roc(as.numeric(y_test == levels(y_test)[1]), model1_prob[, levels(y_test)[1]])
model1_roc2 <- roc(as.numeric(y_test == levels(y_test)[2]), model1_prob[, levels(y_test)[2]])
model1_roc3 <- roc(as.numeric(y_test == levels(y_test)[3]), model1_prob[, levels(y_test)[3]])

# Plot ROC
plot(model1_roc1, col = cols[1], main = "ROC - Multinomial Logistic Regression")
lines(model1_roc2, col = cols[2])
lines(model1_roc3, col = cols[3])

# Compute class AUCs
model1_class_aucs <- sapply(levels(y_test), function(cl) round(auc(roc(as.numeric(y_test == cl), model1_prob[, cl])), 3))

# Add legend with class AUCs + macro AUC
legend("bottomright",
       legend = c(paste0("Class ", levels(y_test), " (AUC=", model1_class_aucs, ")"),
                  paste0("Macro AUC=", round(model1__macro_auc, 3)),
                  paste0("Accuracy=", round(model1_accuracy, 3)),
                  paste0("Macro F1 Score=", round(model1_macro_f1, 3))),
       col = c(cols, rep(NA, 3)),  
       lwd = c(rep(2, 3), rep(NA, 3)))
```

### 1.2 Confusion matrix heatmap
```{r}
model1_cm <- table(True = y_test, Pred = model1_pred)
model1_cm_df <- as.data.frame(model1_cm) |>
  rename(True = True, Predicted = Pred, Count = Freq)

ggplot(model1_cm_df, aes(x = Predicted, y = True, fill = Count)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Count), color = "black", size = 5) +
  scale_fill_gradient(low = "white", high = "purple") +
  labs(title = "Confusion Matrix Heatmap - Multi Log Regression", x = "Predicted", y = "True") +
  theme_minimal(base_size = 14)
```



## Model 2 Preparation
```{r}
df_ranger <- diabetes |>
  select(diabetes_2, high_bp, high_chol, chol_check, bmi, smoker, stroke, 
         heart_diseaseor_attack, phys_activity, fruits, veggies, hvy_alcohol_consump,
         any_healthcare, no_docbc_cost, gen_hlth, ment_hlth, phys_hlth, diff_walk, 
         sex, age, education, income)

df_ranger$diabetes_2 <- factor(df_ranger$diabetes_2, 
                              levels = c(0, 1), 
                              labels = c("No_Diabetes", "Diabetes"))
df_ranger
```

## Model 2: Binary Ranger/RandomForest Model
### 2.1 Data splitting
```{r}
set.seed(1) 

index <- createDataPartition(df_ranger$diabetes_2, p = 0.8, list = FALSE)
train_data <- df_ranger[index, ]
test_data <- df_ranger[-index, ]

y_test <- test_data$diabetes_2
```


### 2.2 Train Binary Random Forest Model (using ranger)
```{r}
rf_model <- ranger(
  dependent.variable.name = "diabetes_2", 
  data = train_data,
  num.trees = 500,
  importance = "impurity",
  seed = 1,
  verbose = TRUE,
  probability = TRUE
)
```

### 2.3 Prediction and Metrics
```{r}
pred <- predict(rf_model, data = test_data, type = "response")

# Probability of Diabetes (positive class)
model_prob <- pred$predictions[, "Diabetes"]

# Class predictions = class with highest probability
model_pred <- colnames(pred$predictions)[ max.col(pred$predictions) ]
model_pred <- factor(model_pred, levels = levels(y_test))

# Confusion matrix
cm <- confusionMatrix(model_pred, y_test)

# Accuracy
model_accuracy <- as.numeric(cm$overall['Accuracy'])

# F1 for Diabetes
f1_diabetes <- as.numeric(cm$byClass['F1'])
# F1 for No_Diabetes
cm_rev <- confusionMatrix(model_pred, y_test, positive = "No_Diabetes")
f1_no_diabetes <- as.numeric(cm_rev$byClass['F1'])
# Macro F1
model_macro_f1 <- mean(c(f1_diabetes, f1_no_diabetes))

# AUC for Diabetes
model_roc <- roc(response = y_test, predictor = model_prob)
model_auc <- as.numeric(model_roc$auc)

cat("Accuracy is:     ", round(model_accuracy, 4), "\n")
cat("Macro F1 score is:", round(model_macro_f1, 4), "\n")
cat("ROC AUC (Binary) is:", round(model_auc, 4), "\n")
```

### 2.4 Confusion matrix heatmap
```{r}
model_cm_table <- cm$table

model_cm_df <- as.data.frame(model_cm_table) |>
  rename(True = Reference, Predicted = Prediction, Count = Freq)

p_heatmap <- ggplot(model_cm_df, aes(x = Predicted, y = True, fill = Count)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Count), color = "black", size = 5) +
  scale_fill_gradient(low = "#e0f2f1", high = "#00796b") +
  labs(
    title = "Confusion Matrix Heatmap - Random Forest", 
    x = "Predicted Class", 
    y = "True Class"
  ) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(hjust = 0.5))

print(p_heatmap)
```

### 2.5 ROC curve
```{r}
plot(model_roc, 
     col = "darkred", 
     main = "ROC Curve - Binary Random Forest", 
     lwd = 2)

abline(a = 0, b = 1, lty = 2, col = "grey50") 

legend(
  "bottomright",
  legend = c(
    paste0("Diabetes vs No Diabetes (AUC=", round(model_auc, 3), ")"),
    paste0("Accuracy=", round(model_accuracy, 3)),
    paste0("Macro F1 Score=", round(model_macro_f1, 3))
  ),
  col = c("darkred", rep(NA, 2)),
  lwd = c(2, rep(NA, 2))
)
```

### 2.6 Variable importance
```{r}
importance_df <- data.frame(
  Variable = names(rf_model$variable.importance),
  Importance = rf_model$variable.importance
) |>
  arrange(desc(Importance))

ggplot(importance_df[1:15, ], aes(x = reorder(Variable, Importance), y = Importance)) +
  geom_col(fill = "#00796b") +
  coord_flip() +
  labs(title = "Top 15 Variable Importances (Random Forest)",
       x = "Variable",
       y = "Importance") +
  theme_minimal(base_size = 14)

```

## Model 3: Binary Logistic Regression Model

```{r}
diabetes_temp = diabetes %>%
  mutate(
    diabetes_binary = ifelse(diabetes_012 == "No Diabetes", 0, 1),
    diabetes_binary = as.numeric(diabetes_binary)
  )
```

We fit a logistic regression model using health and lifestyle predictors such as blood pressure, cholesterol, BMI, smoking status, physical activity, age, and sex.

### 3.1 Fit the Model

```{r}
logit_model = glm(
  diabetes_binary ~ high_bp + high_chol + bmi + smoker + phys_activity + age + sex,
  data = diabetes_temp,
  family = binomial(link = "logit")
)

summary(logit_model)

```

### 3.2 Odds Ratios and 95% Confidence Intervals

```{r}

exp(cbind(
  OR = coef(logit_model),
  confint.default(logit_model)
))
```

### 3.3 Predicted Probability, Class Prediction, and Confusion Matrix

```{r}
# Predicted probability
diabetes_temp$pred_prob <- predict(logit_model, type = "response")

# Convert to class prediction at threshold = 0.5
diabetes_temp$pred_class <- ifelse(diabetes_temp$pred_prob > 0.5, 1, 0)

# Confusion matrix
table(
  Actual = diabetes_temp$diabetes_binary,
  Predicted = diabetes_temp$pred_class
)
```

### 3.4 ROC Curve and AUC

```{r}
library(pROC)

roc_obj <- roc(diabetes_temp$diabetes_binary, diabetes_temp$pred_prob)
auc(roc_obj)

plot(roc_obj, main = "ROC Curve for Logistic Regression")
```

## Model 4: Binary XGBoost 

We trained an XGBoost model using a 70/30 train-test split. Predictors were converted to numeric matrices using model.matrix.

### 4.1 Train/Test Split and Data Preparation

```{r}
library(xgboost)
library(caret)
library(dplyr)

#Train/Test Split
set.seed(123)

train_index = createDataPartition(diabetes_temp$diabetes_binary, p = 0.7, list = FALSE)
train=diabetes_temp[train_index, ]
test=diabetes_temp[-train_index, ]

train$binary_outcome=as.numeric(train$diabetes_binary)
test$binary_outcome=as.numeric(test$diabetes_binary)

x_train <- model.matrix(diabetes_binary ~ . - 1, data = train)
x_test  <- model.matrix(diabetes_binary ~ . - 1, data = test)

dtrain=xgb.DMatrix(data = as.matrix(x_train), label = train$diabetes_binary)
dtest=xgb.DMatrix(data = as.matrix(x_test),  label = test$diabetes_binary)
```

### 4.2 Train the XGBoost Model

```{r}
# Set Hyperparameters

params = list(
  objective = "binary:logistic",
  eval_metric = "auc",
  max_depth = 4,
  eta = 0.1,
  subsample = 0.8,
  colsample_bytree = 0.8
)
```

```{r}
# Train Model

xgb_model = xgb.train(
  params = params,
  data = dtrain,
  nrounds = 200,
  watchlist = list(train = dtrain, test = dtest),
  verbose = 0
)
```


### 4.3 Predictions and Confusion Matrix
```{r}
pred_prob <- predict(xgb_model, dtest)
pred_class <- ifelse(pred_prob >= 0.5, 1, 0)

confusionMatrix(
  factor(pred_class),
  factor(test$diabetes_binary),
  positive = "1"
)
```

### 4.4 AUC

```{r}
xgb_auc = roc(test$diabetes_binary, pred_prob)
auc(xgb_auc)
```

### 4.5 Variable Importance Plot

```{r}

importance_matrix <- xgb.importance(model = xgb_model)
xgb.plot.importance(importance_matrix)
```