---
title: "Exploring Diabetes Risk Factors and Predictive Modeling in the United States"
output:
  html_document:
    self_contained: false
    code_folding: hide
    toc: true
    toc_float: true
    theme: cosmo

---

*Yvonne Xie, Rosie Kwon, Vicky Yang, Yiying Chen, Charlotte Zhang*


## Introduction
```{r child="introduction.Rmd"}
```

## Data
```{r child="Data.Rmd"}
```

## Exploratory Data Analysis

```{r, include = FALSE, message = FALSE, warning = FALSE}
# Load required libraries
library(tidyverse)
library(rvest)
library(plotly)

diabetes <- readRDS("clean_diabetes_data.rds")
  
binary_vars = c(
  "high_bp", "high_chol", "chol_check",
  "smoker", "stroke", "heart_diseaseor_attack",
  "phys_activity", "hvy_alcohol_consump", "any_healthcare",
  "no_docbc_cost", "diff_walk", "fruits", "veggies"
)
```


## <b>Diabetes Status Overview </b>

To begin the analysis, we examine how diabetes variable is distributed across the three diagnostic categories.

Because prediabetes accounts for a very small proportion of the sample, and because both diabetes and prediabetes represent elevated metabolic risk, we combined these two categories into a single “at-risk” group. 
```{r}
diabetes =
  diabetes |> 
  mutate(
    diabetes_binary = 
      factor(diabetes_2,
             levels = c(0,1),
             labels = c("No Diabetes", "Diabetes")),
  )

diabetes |>  
  count(diabetes_binary) |>  
  mutate(
    percent = round(n / sum(n) * 100, 1)
  ) |> 
  ggplot(aes(x = diabetes_binary, y = n, fill = diabetes_binary)) +
  geom_col() +
  geom_text(aes(label = paste0(percent, "%"),vjust = -0.2), 
  ) +
  scale_fill_manual(
  values = c(
    "No Diabetes" = "skyblue",
    "Diabetes" = "tomato"
    )
  ) +
  labs(
    title = "Diabetes distribution",
    x = "Diabetes Category",
    y = "Count"
  ) +
  theme_minimal()
```

The 84.2% of population in the dataset do not have diabetes while 15.6% reported having Diabetes. Although diabetes affects a smaller portion of the sample, it still represents a substantial public health concern given the large population size.


## <b>Demographic Characteristics & Diabetes Patterns </b>

We examined the demographic structure of the BRFSS sample dataset, including age, sex, education, income, and general health.

Understanding population characteristics provides important context for interpreting diabetes prevalence and identifying meaningful risk patterns.

### 1) Sex & Age

```{r error = FALSE}
diabetes |>
  group_by(age, sex) |> 
  summarize(
    diabetes_rate = mean(diabetes_2 == 1)) |>  
  plot_ly(
    x = ~age,
    y = ~diabetes_rate,
    color = ~sex,
    colors = c("tomato","skyblue"),
    type = "scatter",
    mode = "markers"
  ) |>  
  layout(
    title = "Diabetes prevalence by Sex and Age Group",
    xaxis = list(title = "Age Group"),
    yaxis = list(title = "Diabetes Rate")
  )
```

Diabetes prevalence increases steadily with age for both males and females. The rate of diabetes increases rapidly from 40s and peak at aged 70-74. Prevalence of diabetes is higher in male compared to female across all age groups, indicating that a sex-specific pattern in diabetes risk that becomes more pronounced with age. 

### 2) Education 

In order to compare diabetes risk across education levels, a proportional plot was used since the large differences in group sizes can mask prevalence differences. 

```{r}
ggplot(diabetes, aes(y = education, fill = diabetes_binary)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c(
    "No Diabetes" = "lightgreen",
    "Diabetes" = "salmon"
  )) +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(
    title = "Proportion of Diabetes Status by Education Level",
    x = "Proportion",
    y = "Education Level",
    fill = "Diabetes Status"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(hjust = 1))
```

The proportional plot indicates that the distribution of diabetes status is relatively consistent across education levels. There is  gradient suggesting that lower educational attainment is slightly more associated with markedly different diabetes prevalence.

### 3) Income

Even though the number of diabetes cases appears visually similar across income levels, we used a proportional plot for income variable as there are large differences in total sample size between income groups..

```{r}
ggplot(diabetes, aes(y = income, fill = diabetes_binary))+
  geom_bar(position = "fill") +
  scale_fill_manual(values = c(
    "No Diabetes" = "lightgreen",
    "Diabetes" = "salmon"
  )) +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(
    title = "Proportion of Diabetes Status by Income Level",
    x = "Proportion",
    y = "Income Level",
    fill = "Diabetes Status"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(hjust = 1))
```

The proportional plot indicates an inverse relationship between income and diabetes prevalence. Individuals with lower income levels show a higher proportion of diabetes, whereas those with higher income levels have a noticeably lower diabetes proportion.

## Characteristics analysis
To evaluate whether key variables differ significantly between diabetes and non-diabetes groups, chi-square tests were conducted for categorical variables and t-tests were performed for continuous variables. In this analysis, the binary diabetes dataset was used, where prediabetes and diabetes were combined into a single diabetes group. Among the 21 variables in the dataset, 17 are categorical and 4 are continuous. A chi-square test for categorical variables assesses whether the distribution of a characteristic differs significantly between groups, whereas a t-test for continuous variables compares group means to determine whether differences in average values are statistically significant.

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(knitr)
chi_diabetes = read.csv("Diabete datasets/diabetes_binary_health_indicators_BRFSS2015.csv")

cat_vars <- c(
  "HighBP", "HighChol", "CholCheck","Smoker", "Stroke", "HeartDiseaseorAttack",
  "PhysActivity", "Fruits", "Veggies", "HvyAlcoholConsump",
  "AnyHealthcare", "NoDocbcCost", "DiffWalk",
  "Sex", "Age", "Education", "Income"
)

chi_results <- list()

for (v in cat_vars) {
  tbl <- table(chi_diabetes$Diabetes_binary, chi_diabetes[[v]])
  test <- chisq.test(tbl)
  chi_results[[v]] <- test
}

chi_table <- data.frame(
  Variable   = names(chi_results),
  Chi_square = sapply(chi_results, function(x) unname(x$statistic)),
  df         = sapply(chi_results, function(x) unname(x$parameter)),
  p_value    = sapply(chi_results, function(x) x$p.value)
)

chi_table$p_value <- format(chi_table$p_value, scientific = TRUE) #show detailed p-value result

kable(chi_table, digits = 4, row.names = FALSE, caption = "Chi-square test results for categorical variables") #put everything neatly in a table
```

Based on the **chi-square test results**, all categorical variables show extremely small p-values. Specifically, HighBP, HighChol, Stroke, HeartDiseaseorAttack, PhysActivity, DiffWalk, Age, Education, and Income have p-values effectively equal to zero, indicating that these characteristics differ substantially between the diabetes and non-diabetes groups. The chi-square statistics are also very large, suggesting strong deviations from what would be expected if the two variables were independent. In other words, these health and demographic characteristics are strongly associated with diabetes status in this dataset.

```{r t_test, warning=FALSE, message=FALSE}
cont_vars <- c("BMI", "MentHlth", "PhysHlth", "GenHlth")

# Run t-tests and extract key results
t_results <- lapply(cont_vars, function(v) {
  
  tt <- t.test(chi_diabetes[[v]] ~ chi_diabetes$Diabetes_binary)
  
  data.frame(
    Variable        = v,
    Mean_non_diab   = unname(tt$estimate[1]),
    Mean_diab       = unname(tt$estimate[2]),
    t_statistic     = unname(tt$statistic),
    df              = unname(tt$parameter),
    p_value         = tt$p.value
  )
})

# Combine list rows into a data frame
t_table <- bind_rows(t_results)

# Format values
t_table <- t_table |>
  mutate(
    Mean_non_diab = round(Mean_non_diab, 2),
    Mean_diab     = round(Mean_diab, 2),
    t_statistic   = round(t_statistic, 2),
    df            = round(df, 1),
    p_value       = p_value)

t_table$p_value <- formatC(t_table$p_value, format = "e", digits = 2)

# Print table
kable(t_table, caption = "Summary of t-tests")
```
```{r t_test_plot, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)

df_cont <- 
  chi_diabetes |>
  pivot_longer(cols = all_of(cont_vars),
               names_to = "variable",
               values_to = "value")

ggplot(df_cont,aes(x = factor(Diabetes_binary,
                    levels = c(0, 1),
                    labels = c("Non-diabetes", "Diabetes")),
                    y = value,
                    fill = factor(Diabetes_binary))) +
  geom_boxplot(outlier.alpha = 0.15) +
  facet_wrap(~ variable, scales = "free_y") +
  labs(
    title = "Continuous variables of Diabetes Status",
    x = "Diabetes Status",
    y = "Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

Similarly, the **t-test results** show that all continuous variables have extremely small p-values, indicating significant mean differences between individuals with and without diabetes. Specifically, participants in the diabetes group have higher BMI, report more days of poor physical and mental health, and assign worse general health ratings compared with non-diabetic individuals. These substantial mean differences, together with large t-statistics and narrow confidence intervals, provide strong evidence that these continuous health measures are associated with diabetes status.

### Discussion: 
 
In general, all characteristics show significant differences between individuals with and without diabetes. This aligns with expectations, as many health and demographic factors, such as blood pressure, cholesterol, physical activity, age, and income, are known to be associated with diabetes risk. One thing that we didn’t expect is that many p-values are extremely small, indicating exceptionally strong group differences. This suggests that the predictors in this dataset have very strong associations with diabetes status, much stronger than what might typically be observed in population-level studies. Overall, these results highlight clear disparities in health behaviors, health status, and demographic characteristics between diabetic and non-diabetic individuals, providing insight into the underlying factors that may contribute to diabetes risk.
